Variability in upwelling patterns across the Benguela

### 2019-10-15 Write_up/Paper1_writeup.Rmd
updated

In this equation, μ represents the wind speed (m/s) and θ represents the wind direction, which is measured in degrees, and the coastal angle as the angle of each of the sites perpendicular to the coastline. The index relies heavily on wind speed and direction data in order to determine the intensity of upwelling and the likelihood of the phenomenon occurring. The above equation produces a value called the upwelling index; for this study specifically, upwelling was classified to occur when the index is greater than 1. However to further ascertain a threshold for upwelling, the `exceedance()` function within the _`heatwaveR`_ package [@schlegel2017heat] was utilised, specifically to determine the consecutive number of days that the event must exceed for it to be regarded and classified as an upwelling event. However it should be noted that upwelling is known to vary on a seasonal basis and also may occur hourly. Therefore this study sought to establish the minimum duration for the analyses of upwelling as 1 day; the rationale behind this being that data from the SACTN dataset as well as the Satellite remotely sensed SST data is collected daily. Furthermore, it was also determined that a 25th percentile threshold for the upwelling index result lasting for >= 1 days matched with the upwelling index value of 1. As a result the 25th percentile threshold was used as a more acceptable value to predict upwelling events. In order to determine the upwelling metrics, the `detect_event()` function within the _`heatwaveR`_ package was used.An R script (https://github.com/AmierohAbrahams/Upwelling_MCS) was used to calculate the individual upwelling events for both the *in situ* and remotely sensed SST data. Because upwelling events were calculated relative to percentile exceedances, rather than an absolute definition such as periods with temperatures above a fixed temperature threshold, upwelling events occurred any time of the year with upwelling shown to be more dominant during spring and summer months. It must also be stated that due to irregular sampling efforts in the SACTN dataset the spatial and temporal distributions of upwelling is not necessarily even.

@@

Sea Point showcased the most intense upwelling event at 10 km from the coastline during the year 2013, lasting for 21 days (Refer to figure, do this throughout). The majority of upwelling events in Sea Pont occurred around 10 km away from the coastline lasted for less than five days, and had an approximate cumulative intensity ranging between of -10.00deg. C x days and 12.00 deg. C x days. Conversely, at shorter distances like 30 km from the coastline several upwelling events lasted for longer than 15 days with a cumulative intensity ranging between -30deg. C x days and -20deg. C x days. At a distance of 30 km both dominant events appear at a distance of 10 km are present along with several other additional events (This doesn’t make sense). At a distance of 40 km, less intense events occurred. At a distance of 50000 km from the coastline, several events appear to have lasted for approximately 2 – 3 days with the most intense events still occurring in 2015.Events within Saldanha Bay typically lasted for approximately three days but varying at different distances. At a distance of 10 km one major event occurred that lasted for approximately 27 days with a cumulative intensity of -61.70deg. C x days. This event was also present at a distance of 20 km from the coastline, albeit at a lesser intensity, but was not present at 30km, 40km and 50km respectively. Events occurring within 30 km and 40 km from the coastline lasting for approximately 10 days in most cases. However, the majority of events were observed at a distance of 10 km and 20 km from the coastline. Fewer events occurred at a distance of 50 km, with most of the events occurring at this distance having lasted for less than five days.

@@

Within the SACTN dataset, the majority of upwelling events detected in Lamberts Bay lasted for longer than five days. The longest upwelling event was detected in 2012 lasting for approximately 12 days with a cumulative intensity of -12.63deg. C x days. Several upwelling events were detected at this site. Only eight upwelling events were recorded and detected at Port Nolloth. Here most of the lasted between three to five days with the most intense event lasting for 11 days with a cumulative intensity of -12.81deg. C x days. Saldanha Bay upwelling events also lasted between three to five days. The event with the longest duration occurred in 2016 with a cumulative intensity of -22.58deg. C x days. Events observed within Sea Point lasted less than five days, but two events lasted for 21 days. Those events represented intensities of -54.54 and -36.81 deg. C x days respectively.

@@

CMC observations of upwelling events at Sea Point detected durations of less than four days. The most intense event occurred in 2010, and I found that intensifies appeared to increase with distance from the coastline. The fewest number of events occurred 50 km away from the coastline. At Saldanha Bay, the duration of upwelling events mostly ranged between three to four days. Similar intensity patterns occurred at 10 km, 20 km and 30km from the coastline respectively. At 40 km an event representing a cumulative intensity of -28.64deg. C x days for was detected for 20 days within 2010. At Port Nolloth, most of the events are found to occur for duration of three days. Two major upwelling events were present at different distances within 2010 with a variation in their respective cumulative intensities. Those events both lasted for approximately seven days. Similar patterns were found within Lamberts Bay. At 30 km a major event was detected that lasted for approximately 19 days, but most events here lasted for less than five days. At a distance of 40 km and 50 km from the coastline similar patterns were observed, with most events being detected at 40 km and 50 km respectively.

@@

Results gathered during the course of this study clearly illustrate that upwelling events detected in different ocean sea surface temperature datasets showed variations with regards to duration and intensities across different time intervals. Noticeably, some patterns were dominant and persistent throughout the datasets; however, it was also observed that fewer upwelling events occurred at distances furthest away from the coastline. This was correlated in satellite datasets, where, often at distances of 10km and 20km from the coastline, similar upwelling events were apparent. This difference appears to be related to nature and the variability of physical oceanographic processes at both the broader and local scales; however, the potential involvement of other processes cannot be ruled out. These findings illustrate how dis-similar satellite data is from in situ collected datasets.

>>>>

In this equation, μ represents the wind speed (m/s) and θ represents the wind direction, which is measured in degrees, and the coastal angle as the angle of each of the sites perpendicular to the coastline. The index relies heavily on wind speed and direction data in order to determine the intensity of upwelling and the likelihood of the phenomenon occurring. The above equation produces a value called the upwelling index; for this study specifically, upwelling was classified to occur when the index value is greater than 1. However to further ascertain a threshold for upwelling, the `exceedance()` function within the _`heatwaveR`_ package [@schlegel2017heat] was utilised, specifically to determine the consecutive number of days that the event must exceed for it to be regarded and classified as an upwelling event. However it should be noted that upwelling is known to vary on a seasonal basis and also may occur hourly. Therefore this study sought to establish the minimum duration for the analyses of upwelling as 1 day; the rationale behind this being that data from the SACTN dataset as well as the satellite remotely sensed SST data is collected daily. Furthermore, it was also decided to make use of a percentile value rather than a static upwelling index value of >1 as percentile values are often more approved. It was then determined that a 25th percentile threshold for the upwelling index result lasting for >= 1 days matched with the upwelling index value of >1. As a result the 25th percentile threshold was used as a more acceptable value to predict upwelling events. In order to determine the upwelling metrics, the `detect_event()` function within the _`heatwaveR`_ package was used.An R script (https://github.com/AmierohAbrahams/Upwelling_MCS) was used to calculate the individual upwelling events for both the *in situ* and remotely sensed SST data. Because upwelling events were calculated relative to percentile exceedances, rather than an absolute definition such as periods with temperatures above a fixed temperature threshold, upwelling events occurred any time of the year with upwelling shown to be more dominant during spring and summer months. It must also be stated that due to irregular sampling efforts in the SACTN dataset the spatial and temporal distributions of upwelling is not necessarily even.

@@

Sea Point showcased the most intense upwelling event at 10 km from the coastline during the year 2013, lasting for 21 days (Fig 2). The majority of upwelling events in Sea Pont occurred around 10 km away from the coastline lasted for less than five days, and had an approximate cumulative intensity ranging between of -10.00deg. C x days and 12.00 deg. C x days. Conversely, at shorter distances like 30 km from the coastline several upwelling events lasted for longer than 15 days with a cumulative intensity ranging between -30deg. C x days and -20deg. C x days. At a distance of 30 km both dominant events appear at a distance of 10 km are present along with several other additional events (This doesn’t make sense). At a distance of 40 km, less intense events occurred. At a distance of 50000 km from the coastline, several events appear to have lasted for approximately 2 – 3 days with the most intense events still occurring in 2015.Events within Saldanha Bay typically lasted for approximately three days but varying at different distances. At a distance of 10 km one major event occurred that lasted for approximately 27 days with a cumulative intensity of -61.70deg. C x days. This event was also present at a distance of 20 km from the coastline, albeit at a lesser intensity, but was not present at 30km, 40km and 50km respectively. Events occurring within 30 km and 40 km from the coastline lasting for approximately 10 days in most cases. However, the majority of events were observed at a distance of 10 km and 20 km from the coastline. Fewer events occurred at a distance of 50 km, with most of the events occurring at this distance having lasted for less than five days.

@@

Within the SACTN dataset (Fig 3), the majority of upwelling events detected in Lamberts Bay lasted for longer than five days. The longest upwelling event was detected in 2012 lasting for approximately 12 days with a cumulative intensity of -12.63deg. C x days. Several upwelling events were detected at this site. Only eight upwelling events were recorded and detected at Port Nolloth. Here most of the lasted between three to five days with the most intense event lasting for 11 days with a cumulative intensity of -12.81deg. C x days. Saldanha Bay upwelling events also lasted between three to five days. The event with the longest duration occurred in 2016 with a cumulative intensity of -22.58deg. C x days. Events observed within Sea Point lasted less than five days, but two events lasted for 21 days. Those events represented intensities of -54.54 and -36.81 deg. C x days respectively.

@@

CMC observations of upwelling events (Fig 4) at Sea Point detected durations of less than four days. The most intense event occurred in 2010, and I found that intensifies appeared to increase with distance from the coastline. The fewest number of events occurred 50 km away from the coastline. At Saldanha Bay, the duration of upwelling events mostly ranged between three to four days. Similar intensity patterns occurred at 10 km, 20 km and 30km from the coastline respectively. At 40 km an event representing a cumulative intensity of -28.64deg. C x days for was detected for 20 days within 2010. At Port Nolloth, most of the events are found to occur for duration of three days. Two major upwelling events were present at different distances within 2010 with a variation in their respective cumulative intensities. Those events both lasted for approximately seven days. Similar patterns were found within Lamberts Bay. At 30 km a major event was detected that lasted for approximately 19 days, but most events here lasted for less than five days. At a distance of 40 km and 50 km from the coastline similar patterns were observed, with most events being detected at 40 km and 50 km respectively.

@@

Results gathered during the course of this study clearly illustrate that upwelling events detected in different ocean sea surface temperature datasets showed variations with regards to duration and intensities across different distances. Noticeably, some patterns were dominant and persistent throughout the datasets; however, it was also observed that fewer upwelling events occurred at distances furthest away from the coastline. This was correlated in satellite datasets, where, often at distances of 10km and 20km from the coastline, similar upwelling events were apparent. This difference appears to be related to nature and the variability of physical oceanographic processes at both the broader and local scales; however, the potential involvement of other processes cannot be ruled out. These findings illustrate how dis-similar satellite data is from in situ collected datasets.
